---
title: "library-aligner benchmarking"
author: "Oscar G Wilkins"
format: html
editor: visual
---

```{r setup, message=FALSE, warning=FALSE}
library(reticulate)
use_condaenv("library_aligner", conda = "/Users/ogw/miniconda3/bin/conda")
```

## Configuration

```{python config}
import os, gzip, random, subprocess, time
import psutil, pysam, dnaio, pandas as pd

LA_PYTHON = "/Users/ogw/miniconda3/envs/library_aligner/bin/python"
LA_SCRIPT = "/Users/ogw/Documents/GitHub/library-aligner/src/library_aligner/core.py"
OUT_ROOT  = "/tmp/la_bench"

WILDCARD = "NNNNNNNNNNNNNNN"   # 15 bp wildcard
F5       = "ACGTACGTAC"        # 10 bp left flank
F3       = "TGCATGCATG"        # 10 bp right flank
BASES    = ['A', 'C', 'G', 'T']

# Maximum reads sent to each aligner per condition.
# Keeping this low makes naive minimap2 tractable at large library sizes
# while still giving a fair comparison. Change here to affect all experiments.
MAX_READS = 10

print("Config OK")
```

## Helper Functions

```{python helpers}
# ── Sequence utilities ────────────────────────────────────────────────────────

def mutate_seq(seq, error_rate, rng):
    """Introduce substitutions at a given rate."""
    return "".join(
        rng.choice([b for b in BASES if b != c]) if rng.random() < error_rate else c
        for c in seq
    )


# ── Benchmark data generators ─────────────────────────────────────────────────

def generate_benchmark_data(
    out_dir, num_barcodes, read_depth=100, seed=42,
    backbone_error_rate=0.05, backbone_length=400,
):
    """Generate wildcard FASTA, linkage CSV, and a noiseless FASTQ."""
    rng = random.Random(seed)
    os.makedirs(out_dir, exist_ok=True)
    fasta_path = os.path.join(out_dir, "reference.fasta")
    csv_path   = os.path.join(out_dir, "linkage.csv")
    fastq_path = os.path.join(out_dir, "reads.fastq.gz")

    backbone_l = "".join(rng.choices(BASES, k=backbone_length))
    backbone_r = "".join(rng.choices(BASES, k=backbone_length))
    barcodes   = ["".join(rng.choices(BASES, k=15)) for _ in range(num_barcodes)]

    bc_templates = {}
    for i, bc in enumerate(barcodes):
        mut_l    = mutate_seq(backbone_l, backbone_error_rate, rng)
        mut_r    = mutate_seq(backbone_r, backbone_error_rate, rng)
        template = mut_l + F5 + WILDCARD + F3 + mut_r
        bc_templates[bc] = (f"Ref_{i}", template)

    with open(fasta_path, "w") as fa, open(csv_path, "w") as csv_f:
        csv_f.write("barcode,rname\n")
        for bc, (rname, template) in bc_templates.items():
            fa.write(f">{rname}\n{template}\n")
            csv_f.write(f"{bc},{rname}\n")

    ground_truth = {}
    with gzip.open(fastq_path, "wt") as fq:
        for i, (bc, (rname, template)) in enumerate(bc_templates.items()):
            resolved = template.replace(WILDCARD, bc)
            for j in range(read_depth):
                qname = f"read_{i}_{j}"
                ground_truth[qname] = bc
                fq.write(f"@{qname}\n{resolved}\n+\n{'I' * len(resolved)}\n")

    return fasta_path, csv_path, fastq_path, barcodes, ground_truth


def generate_naive_fasta(out_dir, fasta_path, csv_path):
    """
    Build a concatenated FASTA where every barcode gets its own entry
    (wildcard replaced with the actual barcode). Entry names ARE barcodes —
    this is both the naive alignment reference and the simulation source.
    """
    naive_fasta_path = os.path.join(out_dir, "naive_reference.fasta")
    references, current_name = {}, None
    with open(fasta_path) as f:
        for line in f:
            line = line.strip()
            if line.startswith(">"):
                current_name = line[1:]
                references[current_name] = ""
            else:
                references[current_name] += line

    df = pd.read_csv(csv_path)
    with open(naive_fasta_path, "w") as fa:
        for _, row in df.iterrows():
            bc       = row["barcode"]
            template = references[row["rname"]]
            fa.write(f">{bc}\n{template.replace(WILDCARD, bc)}\n")
    return naive_fasta_path


def generate_spliced_benchmark(
    out_dir, num_barcodes, read_depth=100, seed=42,
    backbone_error_rate=0.05,
    exon_lengths=(200, 200, 200),
    intron_length=200,
):
    """
    Generate benchmark data for spliced alignment.

    Reference FASTA = pre-mRNA (exons + canonical GT-AG introns).
    Reads          = spliced mRNA (exons only, barcode in exon 1).

    This tests that the splice preset correctly handles reads that span
    intron-exon boundaries.
    """
    rng = random.Random(seed)
    os.makedirs(out_dir, exist_ok=True)
    fasta_path = os.path.join(out_dir, "reference.fasta")
    csv_path   = os.path.join(out_dir, "linkage.csv")
    fastq_path = os.path.join(out_dir, "reads.fastq.gz")

    barcodes     = ["".join(rng.choices(BASES, k=15)) for _ in range(num_barcodes)]
    n_exons      = len(exon_lengths)
    ground_truth = {}

    with open(fasta_path, "w") as fa, open(csv_path, "w") as csv_f, \
         gzip.open(fastq_path, "wt") as fq:
        csv_f.write("barcode,rname\n")

        for i, bc in enumerate(barcodes):
            rname = f"Ref_{i}"

            # Per-template exon sequences with divergence
            exons = [
                mutate_seq("".join(rng.choices(BASES, k=l)), backbone_error_rate, rng)
                for l in exon_lengths
            ]

            # Canonical GT-AG introns
            introns = [
                "GT" + "".join(rng.choices(BASES, k=intron_length - 4)) + "AG"
                for _ in range(n_exons - 1)
            ]

            # Barcode placed in exon 1: left + F5 + WILDCARD/bc + F3 + right
            split        = min(50, len(exons[0]))
            left, right  = exons[0][:split], exons[0][split:]
            exon1_wc     = left + F5 + WILDCARD + F3 + right
            exon1_bc     = left + F5 + bc       + F3 + right

            # Genomic reference (wildcard, for library-aligner FASTA)
            genomic = exon1_wc + "".join(
                intron + exon for intron, exon in zip(introns, exons[1:])
            )

            # mRNA read (spliced, actual barcode)
            mrna = exon1_bc + "".join(exons[1:])

            fa.write(f">{rname}\n{genomic}\n")
            csv_f.write(f"{bc},{rname}\n")

            for j in range(read_depth):
                qname = f"read_{i}_{j}"
                ground_truth[qname] = bc
                fq.write(f"@{qname}\n{mrna}\n+\n{'I' * len(mrna)}\n")

    return fasta_path, csv_path, fastq_path, barcodes, ground_truth


# ── Subsampling ───────────────────────────────────────────────────────────────

def subsample_fastq(fastq_path, ground_truth, max_reads, out_dir, seed=42):
    """
    Subsample a FASTQ to at most max_reads reads, guided by the ground truth.
    Returns (new_fastq_path, subsampled_ground_truth).
    If len(ground_truth) <= max_reads the inputs are returned unchanged.
    """
    rng        = random.Random(seed)
    all_qnames = list(ground_truth.keys())

    if len(all_qnames) <= max_reads:
        return fastq_path, ground_truth

    kept     = set(rng.sample(all_qnames, max_reads))
    sub_gt   = {q: ground_truth[q] for q in kept}
    out_path = os.path.join(out_dir, "subsampled.fastq.gz")

    with dnaio.open(fastq_path) as f_in, gzip.open(out_path, "wt") as f_out:
        for record in f_in:
            qname = record.name.split()[0]
            if qname in kept:
                f_out.write(
                    f"@{qname}\n{record.sequence}\n+\n{record.qualities}\n"
                )

    return out_path, sub_gt


# ── Noisy read simulators ─────────────────────────────────────────────────────

def simulate_badread(naive_fasta_path, out_dir, quantity="5x", seed=42):
    """
    Simulate ONT reads with Badread (https://github.com/rrwick/Badread).
    Uses the naive FASTA (entry names = barcodes) so ground truth is free.
    Requires: badread on PATH.
    """
    fastq_path = os.path.join(out_dir, "badread_reads.fastq.gz")
    cmd = [
        "badread", "simulate",
        "--reference",    naive_fasta_path,
        "--quantity",     quantity,
        "--error_model",  "nanopore2023",
        "--qscore_model", "nanopore2023",
        "--seed",         str(seed),
        "--length",       "1000,100",
    ]
    proc = subprocess.run(cmd, capture_output=True)
    with gzip.open(fastq_path, "wt") as out:
        out.write(proc.stdout.decode())
    return fastq_path


def simulate_art(naive_fasta_path, out_dir, read_length=150, fold_coverage=10, seed=42):
    """
    Simulate Illumina reads with ART (https://www.niehs.nih.gov/research/resources/software/biostatistics/art).
    Uses the naive FASTA (entry names = barcodes).
    Requires: art_illumina on PATH.
    """
    prefix     = os.path.join(out_dir, "art_reads")
    cmd = [
        "art_illumina",
        "-ss", "HS25",
        "-i",  naive_fasta_path,
        "-l",  str(read_length),
        "-f",  str(fold_coverage),
        "-o",  prefix,
        "-na",           # no alignment output
        "-rs", str(seed),
        "-q",            # quiet
    ]
    subprocess.run(cmd, capture_output=True)
    art_fq     = prefix + ".fq"
    fastq_path = prefix + ".fastq.gz"
    with open(art_fq) as f_in, gzip.open(fastq_path, "wt") as f_out:
        f_out.write(f_in.read())
    os.remove(art_fq)
    return fastq_path


def parse_gt_badread(fastq_path):
    """
    Parse ground truth from Badread read names.
    Format: '<uuid> <entry_name>,<start>-<end>,<strand> ...'
    Entry name in our naive FASTA IS the barcode, so gt is free.
    """
    gt = {}
    with dnaio.open(fastq_path) as f:
        for record in f:
            qname = record.name.split()[0]
            parts = record.name.split()
            if len(parts) > 1:
                bc = parts[1].split(",")[0]
                gt[qname] = bc
    return gt


def parse_gt_art(fastq_path, known_barcodes):
    """
    Parse ground truth from ART read names.
    Format: '<entry_name>-<readnumber>' where entry_name IS the barcode.
    Sort barcodes by length descending to avoid prefix collisions.
    """
    sorted_bcs = sorted(known_barcodes, key=len, reverse=True)
    gt = {}
    with dnaio.open(fastq_path) as f:
        for record in f:
            qname = record.name.split()[0]
            for bc in sorted_bcs:
                if qname.startswith(bc + "-") or qname.startswith(bc + "/"):
                    gt[qname] = bc
                    break
    return gt


# ── Process monitoring ────────────────────────────────────────────────────────

def _poll_process(proc):
    """Poll a subprocess, returning (elapsed_s, peak_rss_MB)."""
    ps_proc  = psutil.Process(proc.pid)
    peak_mem = 0
    start    = time.time()
    while proc.poll() is None:
        try:
            mem = ps_proc.memory_info().rss
            if mem > peak_mem:
                peak_mem = mem
        except psutil.NoSuchProcess:
            break
        time.sleep(0.05)
    return time.time() - start, peak_mem / 1024**2


# ── Alignment benchmarks ──────────────────────────────────────────────────────

def benchmark_library_aligner(
    fasta_path, csv_path, fastq_path, out_dir,
    preset="splice", kmer=10, w=4, no_cache=False,
):
    tag     = "nocache" if no_cache else "cached"
    out_bam = os.path.join(out_dir, f"la_{tag}.bam")
    cmd = [
        LA_PYTHON, LA_SCRIPT,
        "-fq", fastq_path, "-fa", fasta_path, "-c",  csv_path,
        "-o",  out_bam,    "-wc", WILDCARD,
        "--minimap2_preset", preset,
        "--minimap2_kmer",   str(kmer),
        "--minimap2_w",      str(w),
    ]
    if no_cache:
        cmd.append("--no_mappy_cache")
    elapsed, peak_mem = _poll_process(subprocess.Popen(cmd))
    return elapsed, peak_mem, out_bam


def benchmark_naive(naive_fasta_path, fastq_path, out_dir, preset="splice", kmer=10, w=4):
    out_sam = os.path.join(out_dir, "naive_output.sam")
    cmd = [
        "minimap2", "-ax", preset,
        "-k", str(kmer), "-w", str(w),
        naive_fasta_path, fastq_path,
        "-o", out_sam,
    ]
    elapsed, peak_mem = _poll_process(subprocess.Popen(cmd))
    return elapsed, peak_mem, out_sam


# ── Accuracy measurement ──────────────────────────────────────────────────────

def measure_accuracy_la(bam_path, ground_truth):
    """Check BC:Z tag in library-aligner output against ground truth."""
    total = correct = 0
    with pysam.AlignmentFile(bam_path, "rb") as bam:
        for read in bam.fetch():
            if read.is_secondary or read.is_supplementary:
                continue
            qname = read.query_name
            if qname not in ground_truth:
                continue
            total += 1
            if read.has_tag("BC") and read.get_tag("BC") == ground_truth[qname]:
                correct += 1
    return correct / total if total > 0 else 0.0


def measure_accuracy_naive(sam_path, ground_truth):
    """
    Naive FASTA entry names ARE barcodes, so check whether the reference
    the read mapped to matches the expected barcode in ground_truth.
    """
    total = correct = 0
    with pysam.AlignmentFile(sam_path, "rb") as sam:
        for read in sam.fetch():
            if read.is_secondary or read.is_supplementary or read.is_unmapped:
                continue
            qname = read.query_name
            if qname not in ground_truth:
                continue
            total += 1
            if read.reference_name == ground_truth[qname]:
                correct += 1
    return correct / total if total > 0 else 0.0


def _record_row(vary_key, val, la_t, la_m, la_acc, lac_t, lac_m, lac_acc, nv_t, nv_m, nv_acc):
    return {
        vary_key:                  val,
        "la_time":                 la_t,  "la_mem":                 la_m,  "la_accuracy":                 la_acc,
        "la_nocache_time":         lac_t, "la_nocache_mem":         lac_m, "la_nocache_accuracy":         lac_acc,
        "naive_time":              nv_t,  "naive_mem":              nv_m,  "naive_accuracy":              nv_acc,
    }


# ── Experiment runners ────────────────────────────────────────────────────────

def run_experiment(
    label, scales, fixed_kwargs, vary_key, out_root,
    preset="splice", kmer=10, w=4,
):
    """Standard experiment sweeping vary_key across scales with noiseless reads."""
    rows = []
    for val in scales:
        kwargs  = {**fixed_kwargs, vary_key: val}
        out_dir = os.path.join(out_root, label, str(val))
        print(f"  [{label}] {vary_key}={val} ...")

        fasta, csv, fastq, barcodes, gt = generate_benchmark_data(out_dir, **kwargs)
        naive_fasta                     = generate_naive_fasta(out_dir, fasta, csv)

        # Subsample to MAX_READS for both methods (fair + tractable)
        fastq_sub, gt_sub = subsample_fastq(fastq, gt, MAX_READS, out_dir)

        la_t,  la_m,  la_bam  = benchmark_library_aligner(fasta, csv, fastq_sub, out_dir, preset, kmer, w)
        lac_t, lac_m, lac_bam = benchmark_library_aligner(fasta, csv, fastq_sub, out_dir, preset, kmer, w, no_cache=True)
        nv_t,  nv_m,  nv_sam  = benchmark_naive(naive_fasta, fastq_sub, out_dir, preset, kmer, w)

        la_acc  = measure_accuracy_la(la_bam,  gt_sub)
        lac_acc = measure_accuracy_la(lac_bam, gt_sub)
        nv_acc  = measure_accuracy_naive(nv_sam, gt_sub)

        rows.append(_record_row(vary_key, val, la_t, la_m, la_acc, lac_t, lac_m, lac_acc, nv_t, nv_m, nv_acc))
        print(f"    cached:   {la_t:.1f}s {la_m:.0f}MB acc={la_acc:.3f}")
        print(f"    no-cache: {lac_t:.1f}s {lac_m:.0f}MB acc={lac_acc:.3f}")
        print(f"    naive:    {nv_t:.1f}s {nv_m:.0f}MB acc={nv_acc:.3f}")

    df = pd.DataFrame(rows)
    os.makedirs(os.path.join(out_root, label), exist_ok=True)
    df.to_csv(os.path.join(out_root, label, "results.csv"), index=False)
    return df


def run_spliced_experiment(
    label, scales, fixed_kwargs, vary_key, out_root,
    preset="splice", kmer=10, w=4,
):
    """Experiment with spliced reads aligned to genomic (pre-mRNA) references."""
    rows = []
    for val in scales:
        kwargs  = {**fixed_kwargs, vary_key: val}
        out_dir = os.path.join(out_root, label, str(val))
        print(f"  [{label}] {vary_key}={val} ...")

        fasta, csv, fastq, barcodes, gt = generate_spliced_benchmark(out_dir, **kwargs)
        naive_fasta                     = generate_naive_fasta(out_dir, fasta, csv)

        fastq_sub, gt_sub = subsample_fastq(fastq, gt, MAX_READS, out_dir)

        la_t,  la_m,  la_bam  = benchmark_library_aligner(fasta, csv, fastq_sub, out_dir, preset, kmer, w)
        lac_t, lac_m, lac_bam = benchmark_library_aligner(fasta, csv, fastq_sub, out_dir, preset, kmer, w, no_cache=True)
        nv_t,  nv_m,  nv_sam  = benchmark_naive(naive_fasta, fastq_sub, out_dir, preset, kmer, w)

        la_acc  = measure_accuracy_la(la_bam,  gt_sub)
        lac_acc = measure_accuracy_la(lac_bam, gt_sub)
        nv_acc  = measure_accuracy_naive(nv_sam, gt_sub)

        rows.append(_record_row(vary_key, val, la_t, la_m, la_acc, lac_t, lac_m, lac_acc, nv_t, nv_m, nv_acc))
        print(f"    cached:   {la_t:.1f}s {la_m:.0f}MB acc={la_acc:.3f}")
        print(f"    no-cache: {lac_t:.1f}s {lac_m:.0f}MB acc={lac_acc:.3f}")
        print(f"    naive:    {nv_t:.1f}s {nv_m:.0f}MB acc={nv_acc:.3f}")

    df = pd.DataFrame(rows)
    os.makedirs(os.path.join(out_root, label), exist_ok=True)
    df.to_csv(os.path.join(out_root, label, "results.csv"), index=False)
    return df

def run_noisy_experiment(
    label, scales, fixed_kwargs, vary_key, out_root,
    simulator="badread", preset="splice", kmer=10, w=4,
):
    """
    Experiment with noisy reads simulated by Badread (ONT) or ART (Illumina).
    Ground truth is parsed directly from simulator read names — no manual
    labelling required because naive FASTA entry names ARE barcodes.
    """
    rows = []
    for val in scales:
        kwargs  = {**fixed_kwargs, vary_key: val}
        out_dir = os.path.join(out_root, label, str(val))
        print(f"  [{label}/{simulator}] {vary_key}={val} ...")

        # Generate reference data only (noiseless FASTQ not used here)
        fasta, csv, _, barcodes, _ = generate_benchmark_data(out_dir, **kwargs)
        naive_fasta                = generate_naive_fasta(out_dir, fasta, csv)

        # Simulate noisy reads from naive FASTA (entry names = barcodes)
        if simulator == "badread":
            # Generate ~2x MAX_READS total reads regardless of library size
            target_reads = MAX_READS * 2
            approx_bases = val * 850  # ~850bp per reference
            quantity     = f"{max(1, round((target_reads * 1000) / approx_bases))}x"
            noisy_fastq  = simulate_badread(naive_fasta, out_dir, quantity=quantity)
            gt           = parse_gt_badread(noisy_fastq)
        elif simulator == "art":
            noisy_fastq = simulate_art(naive_fasta, out_dir)
            gt          = parse_gt_art(noisy_fastq, barcodes)
        else:
            raise ValueError(f"Unknown simulator: {simulator}")

        print(f"    {simulator} generated {len(gt)} reads")

        # Subsample to MAX_READS
        fastq_sub, gt_sub = subsample_fastq(noisy_fastq, gt, MAX_READS, out_dir)

        la_t,  la_m,  la_bam  = benchmark_library_aligner(fasta, csv, fastq_sub, out_dir, preset, kmer, w)
        lac_t, lac_m, lac_bam = benchmark_library_aligner(fasta, csv, fastq_sub, out_dir, preset, kmer, w, no_cache=True)
        nv_t,  nv_m,  nv_sam  = benchmark_naive(naive_fasta, fastq_sub, out_dir, preset, kmer, w)

        la_acc  = measure_accuracy_la(la_bam,  gt_sub)
        lac_acc = measure_accuracy_la(lac_bam, gt_sub)
        nv_acc  = measure_accuracy_naive(nv_sam, gt_sub)

        rows.append(_record_row(vary_key, val, la_t, la_m, la_acc, lac_t, lac_m, lac_acc, nv_t, nv_m, nv_acc))
        print(f"    cached:   {la_t:.1f}s {la_m:.0f}MB acc={la_acc:.3f}")
        print(f"    no-cache: {lac_t:.1f}s {lac_m:.0f}MB acc={lac_acc:.3f}")
        print(f"    naive:    {nv_t:.1f}s {nv_m:.0f}MB acc={nv_acc:.3f}")

    df = pd.DataFrame(rows)
    os.makedirs(os.path.join(out_root, label), exist_ok=True)
    df.to_csv(os.path.join(out_root, label, "results.csv"), index=False)
    return df

print("Helper functions defined.")
```

------------------------------------------------------------------------

## Experiment 1 --- Varying Number of Barcodes

Total reads are capped at `MAX_READS` by subsampling. Backbone length and divergence are held constant. This is the key experiment demonstrating library-aligner's scaling advantage.

```{python exp1}
BARCODE_SCALES = [10, 50, 100, 500, 1000, 5000, 10000]
BACKBONE_LEN   = 400
ERROR_RATE     = 0.02

df_exp1 = run_experiment(
    label        = "exp1",
    scales       = BARCODE_SCALES,
    fixed_kwargs = dict(backbone_length=BACKBONE_LEN, backbone_error_rate=ERROR_RATE,
                        read_depth=200),
    vary_key     = "num_barcodes",
    out_root     = OUT_ROOT,
)
print(df_exp1.to_string(index=False))
```

------------------------------------------------------------------------

## Experiment 2 --- Varying Read / Template Length

Number of barcodes and total reads are fixed; backbone length is swept from short (50 bp) to long-read realistic (3000 bp).

```{python exp2}
LENGTH_SCALES = [50, 200, 400, 1000, 2000, 3000]

df_exp2 = run_experiment(
    label        = "exp2",
    scales       = LENGTH_SCALES,
    fixed_kwargs = dict(num_barcodes=250, read_depth=20,
                        backbone_error_rate=ERROR_RATE),
    vary_key     = "backbone_length",
    out_root     = OUT_ROOT,
)
print(df_exp2.to_string(index=False))
```

------------------------------------------------------------------------

## Experiment 3 --- Varying Template Divergence

Number of barcodes and backbone length are fixed; per-base substitution rate is swept from 0 (identical backbones) to 1 (completely random). Tests how sequence similarity between references affects accuracy and speed.

```{python exp3}
DIVERGENCE_SCALES = [0.0, 0.01, 0.05, 0.1, 0.25, 0.5, 1.0]

df_exp3 = run_experiment(
    label        = "exp3",
    scales       = DIVERGENCE_SCALES,
    fixed_kwargs = dict(num_barcodes=500, read_depth=4,
                        backbone_length=BACKBONE_LEN),
    vary_key     = "backbone_error_rate",
    out_root     = OUT_ROOT,
)
print(df_exp3.to_string(index=False))
```

------------------------------------------------------------------------

## Experiment 4 --- ONT Noisy Reads (Badread)

Reads simulated with Badread using the `nanopore2023` error model (\~5--10% error rate). Varies library size identically to Experiment 1 to show that the scaling advantage holds with realistic ONT noise. Ground truth is parsed from Badread read names (no manual labelling).

Note: barcode extraction uses exact matching by default. For heavily degraded barcodes, `--minimum_distance` \> 0 in library-aligner would improve recall.

```{python exp4}
df_exp4 = run_noisy_experiment(
    label        = "exp4_ont",
    scales       = BARCODE_SCALES,
    fixed_kwargs = dict(backbone_length=BACKBONE_LEN, backbone_error_rate=ERROR_RATE,
                        read_depth=20),
    vary_key     = "num_barcodes",
    out_root     = OUT_ROOT,
    simulator    = "badread",
    preset       = "splice",
    kmer         = 10,
    w            = 4,
)
print(df_exp4.to_string(index=False))
```

------------------------------------------------------------------------

## Experiment 5 --- Illumina Reads (ART)

Reads simulated with ART (`HiSeq 2500` profile, 150 bp). Backbone length is reduced to 100 bp so that 150 bp reads reliably span the barcode region. The `splice:sr` preset is used for both library-aligner and naive minimap2.

```{python exp5}
df_exp5 = run_noisy_experiment(
    label        = "exp5_illumina",
    scales       = BARCODE_SCALES,
    fixed_kwargs = dict(backbone_length=100, backbone_error_rate=ERROR_RATE,
                        read_depth=20),
    vary_key     = "num_barcodes",
    out_root     = OUT_ROOT,
    simulator    = "art",
    preset       = "splice:sr",
    kmer         = 10,
    w            = 4,
)
print(df_exp5.to_string(index=False))
```

------------------------------------------------------------------------

## Experiment 6 --- Spliced Reads

Reads are simulated from spliced mRNA (three exons, no introns) but aligned to a genomic reference (pre-mRNA with canonical GT-AG introns). This confirms that the `splice` preset correctly handles junction-spanning reads.

Exon structure: 200 bp \| 200 bp intron \| 200 bp \| 200 bp intron \| 200 bp. Barcode is located in exon 1.

```{python exp6}
df_exp6 = run_spliced_experiment(
    label        = "exp6_spliced",
    scales       = BARCODE_SCALES,
    fixed_kwargs = dict(backbone_error_rate=ERROR_RATE, read_depth=20,
                        exon_lengths=(200, 200, 200), intron_length=200),
    vary_key     = "num_barcodes",
    out_root     = OUT_ROOT,
    preset       = "splice",
    kmer         = 10,
    w            = 4,
)
print(df_exp6.to_string(index=False))
```

------------------------------------------------------------------------

## Plots

```{r load-results, message=FALSE, warning=FALSE}
library(tidyverse)
library(cowplot)

theme_set(
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom")
)

COLS <- c(
  "library-aligner (cached)"   = "#2166ac",
  "library-aligner (no-cache)" = "#f46d43",
  "naive minimap2"             = "#1a9641"
)

pivot_results <- function(df, x_var) {
  df |>
    select(all_of(x_var), starts_with("la_"), starts_with("naive_")) |>
    pivot_longer(
      -all_of(x_var),
      names_to  = c("method", ".value"),
      names_pattern = "^(la_nocache|la|naive)_(.*)"
    ) |>
    mutate(method = recode(method,
      la         = "library-aligner (cached)",
      la_nocache = "library-aligner (no-cache)",
      naive      = "naive minimap2"
    ))
}

make_plots <- function(df, x_var, x_label, log_x = FALSE) {
  d <- pivot_results(df, x_var)

  scale_x <- if (log_x) scale_x_log10() else scale_x_continuous()

  p_time <- ggplot(d, aes(.data[[x_var]], time, colour = method)) +
    geom_line(linewidth = 1) + geom_point(size = 2) +
    scale_colour_manual(values = COLS) + scale_x +
    labs(title = "Run time", x = x_label, y = "Wall time (s)", colour = NULL) +
    ylim(0, NA)

  p_mem <- ggplot(d, aes(.data[[x_var]], mem, colour = method)) +
    geom_line(linewidth = 1) + geom_point(size = 2) +
    scale_colour_manual(values = COLS) + scale_x +
    labs(title = "Peak memory", x = x_label, y = "Peak RSS (MB)", colour = NULL) +
    ylim(0, NA)

  p_acc <- ggplot(d, aes(.data[[x_var]], accuracy, colour = method)) +
    geom_line(linewidth = 1) + geom_point(size = 2) +
    scale_colour_manual(values = COLS) + scale_x +
    scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
    labs(title = "Mapping accuracy", x = x_label, y = "Accuracy", colour = NULL)

  legend_plot <- ggplot(d, aes(.data[[x_var]], time, colour = method)) +
    geom_line() + scale_colour_manual(values = COLS, name = "Method") +
    theme(legend.position = "right")
  legend <- cowplot::get_plot_component(legend_plot, "guide-box-right")

  plots <- plot_grid(
    p_time + theme(legend.position = "none"),
    p_mem  + theme(legend.position = "none"),
    p_acc  + theme(legend.position = "none"),
    ncol = 1
  )
  plot_grid(plots, legend, ncol = 2, rel_widths = c(1, 0.3))
}

exp1 <- read_csv("/tmp/la_bench/exp1/results.csv",         show_col_types = FALSE)
exp2 <- read_csv("/tmp/la_bench/exp2/results.csv",         show_col_types = FALSE)
exp3 <- read_csv("/tmp/la_bench/exp3/results.csv",         show_col_types = FALSE)
exp4 <- read_csv("/tmp/la_bench/exp4_ont/results.csv",     show_col_types = FALSE)
exp5 <- read_csv("/tmp/la_bench/exp5_illumina/results.csv",show_col_types = FALSE)
exp6 <- read_csv("/tmp/la_bench/exp6_spliced/results.csv", show_col_types = FALSE)
```

### Experiment 1 --- Number of Barcodes (noiseless)

```{r plot-exp1, fig.width=10, fig.height=10}
make_plots(exp1, "num_barcodes", "Number of barcodes (log scale)", log_x = TRUE)
```

### Experiment 2 --- Template Length

```{r plot-exp2, fig.width=10, fig.height=10}
make_plots(exp2, "backbone_length", "Backbone length (bp)")
```

### Experiment 3 --- Template Divergence

```{r plot-exp3, fig.width=10, fig.height=10}
make_plots(exp3, "backbone_error_rate", "Per-base substitution rate") +
  scale_x_continuous(labels = scales::percent)
```

### Experiment 4 --- ONT Noisy Reads (Badread)

```{r plot-exp4, fig.width=10, fig.height=10}
make_plots(exp4, "num_barcodes", "Number of barcodes (log scale)", log_x = TRUE)
```

### Experiment 5 --- Illumina Reads (ART)

```{r plot-exp5, fig.width=10, fig.height=10}
make_plots(exp5, "num_barcodes", "Number of barcodes (log scale)", log_x = TRUE)
```

### Experiment 6 --- Spliced Reads

```{r plot-exp6, fig.width=10, fig.height=10}
make_plots(exp6, "num_barcodes", "Number of barcodes (log scale)", log_x = TRUE)
```

------------------------------------------------------------------------

## Session Info

```{r session}
sessionInfo()
```

```{python versions}
import mappy, pysam, edlib
print(f"mappy  {mappy.__version__}")
print(f"pysam  {pysam.__version__}")
print(f"MAX_READS cap: {MAX_READS}")
```
