---
title: "library-aligner benchmarking"
author: "Oscar G Wilkins"
format: html
editor: visual
---

```{r setup, message=FALSE, warning=FALSE}
library(reticulate)
use_condaenv("library_aligner", conda = "/Users/ogw/miniconda3/bin/conda")
```

```{python imports}
import os
import gzip
import random
import subprocess
import time
import psutil
import pandas as pd
import pysam

print("All imports OK")
```

------------------------------------------------------------------------

# Helper Functions

```{python helpers}
LA_PYTHON = "/Users/ogw/miniconda3/envs/library_aligner/bin/python"
LA_SCRIPT = "/Users/ogw/Documents/GitHub/library-aligner/src/library_aligner/core.py"
WILDCARD   = "NNNNNNNNNNNNNNN"  # 15bp
F5         = "ACGTACGTAC"
F3         = "TGCATGCATG"
BASES      = ['A', 'C', 'G', 'T']


def mutate_seq(seq, error_rate, rng):
    """Introduce substitutions at a given rate."""
    return "".join(
        rng.choice([b for b in BASES if b != c]) if rng.random() < error_rate else c
        for c in seq
    )

def generate_benchmark_data(
    out_dir,
    num_barcodes,
    read_depth=100,
    seed=42,
    backbone_error_rate=0.05,
    backbone_length=400,
):
    rng = random.Random(seed)
    os.makedirs(out_dir, exist_ok=True)

    fasta_path = os.path.join(out_dir, "reference.fasta")
    csv_path   = os.path.join(out_dir, "linkage.csv")
    fastq_path = os.path.join(out_dir, "reads.fastq.gz")

    backbone_l = "".join(rng.choices(BASES, k=backbone_length))
    backbone_r = "".join(rng.choices(BASES, k=backbone_length))

    barcodes     = ["".join(rng.choices(BASES, k=15)) for _ in range(num_barcodes)]
    ground_truth = {}

    # Generate one mutated backbone per barcode, shared between fasta and reads
    bc_templates = {}
    for i, bc in enumerate(barcodes):
        mut_l    = mutate_seq(backbone_l, backbone_error_rate, rng)
        mut_r    = mutate_seq(backbone_r, backbone_error_rate, rng)
        template = mut_l + F5 + WILDCARD + F3 + mut_r
        bc_templates[bc] = (f"Ref_{i}", template)

    with open(fasta_path, "w") as fa, open(csv_path, "w") as csv_f:
        csv_f.write("barcode,rname\n")
        for bc, (rname, template) in bc_templates.items():
            fa.write(f">{rname}\n{template}\n")
            csv_f.write(f"{bc},{rname}\n")

    with gzip.open(fastq_path, "wt") as fq:
        for i, (bc, (rname, template)) in enumerate(bc_templates.items()):
            resolved = template.replace(WILDCARD, bc)
            for j in range(read_depth):
                qname = f"read_{i}_{j}"
                ground_truth[qname] = bc
                fq.write(f"@{qname}\n{resolved}\n+\n{'I' * len(resolved)}\n")

    return fasta_path, csv_path, fastq_path, barcodes, ground_truth



def generate_naive_fasta(out_dir, fasta_path, csv_path):
    """
    Build a concatenated multi-reference FASTA where every barcode gets its
    own entry — the naive approach a user would normally take.
    """
    naive_fasta_path = os.path.join(out_dir, "naive_reference.fasta")

    references   = {}
    current_name = None
    with open(fasta_path) as f:
        for line in f:
            line = line.strip()
            if line.startswith(">"):
                current_name = line[1:]
                references[current_name] = ""
            else:
                references[current_name] += line

    df = pd.read_csv(csv_path)
    with open(naive_fasta_path, "w") as fa:
        for _, row in df.iterrows():
            bc       = row["barcode"]
            rname    = row["rname"]
            template = references[rname]
            resolved = template.replace(WILDCARD, bc)
            fa.write(f">{bc}\n{resolved}\n")

    return naive_fasta_path


def _poll_process(proc):
    """Poll a subprocess and return (elapsed_seconds, peak_rss_MB)."""
    ps_proc  = psutil.Process(proc.pid)
    peak_mem = 0
    start    = time.time()
    while proc.poll() is None:
        try:
            mem = ps_proc.memory_info().rss
            if mem > peak_mem:
                peak_mem = mem
        except psutil.NoSuchProcess:
            break
        time.sleep(0.05)
    elapsed = time.time() - start
    return elapsed, peak_mem / 1024**2


def benchmark_library_aligner(
    fasta_path, csv_path, fastq_path, out_dir,
    preset="splice", kmer=10, w=4, no_cache=False
):
    out_bam = os.path.join(out_dir, "la_output.bam")
    cmd = [
        LA_PYTHON, LA_SCRIPT,
        "-fq",  fastq_path,
        "-fa",  fasta_path,
        "-c",   csv_path,
        "-o",   out_bam,
        "-wc",  WILDCARD,
        "--minimap2_preset", preset,
        "--minimap2_kmer",   str(kmer),
        "--minimap2_w",      str(w),
    ]
    if no_cache:
        cmd.append("--no_mappy_cache")
    elapsed, peak_mem = _poll_process(subprocess.Popen(cmd))
    return elapsed, peak_mem, out_bam


def benchmark_naive(naive_fasta_path, fastq_path, out_dir, preset="splice", kmer=10, w=4):
    out_sam = os.path.join(out_dir, "naive_output.sam")
    cmd = [
        "minimap2",
        "-ax", preset,
        "-k",  str(kmer),
        "-w",  str(w),
        naive_fasta_path,
        fastq_path,
        "-o",  out_sam,
    ]
    elapsed, peak_mem = _poll_process(subprocess.Popen(cmd))
    return elapsed, peak_mem, out_sam


def measure_accuracy_la(bam_path, ground_truth):
    """Check BC:Z tag in library-aligner output against ground truth."""
    total = correct = 0
    with pysam.AlignmentFile(bam_path, "rb") as bam:
        for read in bam.fetch():
            if read.is_secondary or read.is_supplementary:
                continue
            qname = read.query_name
            if qname not in ground_truth:
                continue
            total += 1
            if read.has_tag("BC") and read.get_tag("BC") == ground_truth[qname]:
                correct += 1
    return correct / total if total > 0 else 0.0


def measure_accuracy_naive(sam_path, ground_truth):
    """
    For naive approach the FASTA entry name IS the barcode, so check
    whether the reference the read mapped to matches ground truth.
    """
    total = correct = 0
    with pysam.AlignmentFile(sam_path, "rb") as sam:
        for read in sam.fetch():
            if read.is_secondary or read.is_supplementary or read.is_unmapped:
                continue
            qname = read.query_name
            if qname not in ground_truth:
                continue
            total += 1
            if read.reference_name == ground_truth[qname]:
                correct += 1
    return correct / total if total > 0 else 0.0


def run_experiment(label, scales, fixed_kwargs, vary_key, out_root,
                   preset="splice", kmer=10, w=4):
    """
    Generic experiment runner.

    Parameters
    ----------
    label       : str   — experiment name (used for output paths)
    scales      : list  — values of the variable being swept
    fixed_kwargs: dict  — keyword args to generate_benchmark_data that are fixed
    vary_key    : str   — which kwarg in fixed_kwargs is actually the sweep variable
    out_root    : str   — root directory for output files
    """
    rows = []
    for val in scales:
        kwargs = {**fixed_kwargs, vary_key: val}
        print(f"  [{label}] {vary_key}={val} ...")
        out_dir = os.path.join(out_root, label, str(val))

        fasta, csv, fastq, barcodes, gt = generate_benchmark_data(out_dir, **kwargs)
        naive_fasta                     = generate_naive_fasta(out_dir, fasta, csv)

        la_t,  la_m,  la_bam  = benchmark_library_aligner(fasta, csv, fastq, out_dir, preset, kmer, w)
        lac_t, lac_m, lac_bam = benchmark_library_aligner(fasta, csv, fastq, out_dir, preset, kmer, w, no_cache=True)
        nv_t,  nv_m,  nv_sam  = benchmark_naive(naive_fasta, fastq, out_dir, preset, kmer, w)

        la_acc  = measure_accuracy_la(la_bam,  gt)
        lac_acc = measure_accuracy_la(lac_bam, gt)
        nv_acc  = measure_accuracy_naive(nv_sam, gt)

        rows.append({
            vary_key:            val,
            "la_time":           la_t,
            "la_mem":            la_m,
            "la_accuracy":       la_acc,
            "la_nocache_time":   lac_t,
            "la_nocache_mem":    lac_m,
            "la_nocache_accuracy": lac_acc,
            "naive_time":        nv_t,
            "naive_mem":         nv_m,
            "naive_accuracy":    nv_acc,
        })
        print(f"    library-aligner (cached):   {la_t:.1f}s  {la_m:.0f}MB  acc={la_acc:.3f}")
        print(f"    library-aligner (no-cache): {lac_t:.1f}s  {lac_m:.0f}MB  acc={lac_acc:.3f}")
        print(f"    naive:                      {nv_t:.1f}s  {nv_m:.0f}MB  acc={nv_acc:.3f}")

    df = pd.DataFrame(rows)
    os.makedirs(os.path.join(out_root, label), exist_ok=True)
    df.to_csv(os.path.join(out_root, label, "results.csv"), index=False)
    return df


print("Helper functions defined.")
```

------------------------------------------------------------------------

# Experiment 1 --- Varying Number of Barcodes

Total reads are kept fixed at 50,000 by scaling `read_depth` inversely with `num_barcodes`. Backbone length and divergence are held constant.

```{python exp1}
TOTAL_READS    = 10_000
BACKBONE_LEN   = 400
ERROR_RATE     = 0.02
OUT_ROOT       = "/tmp/la_bench"
PRESET         = "splice"
KMER           = 10
W              = 4

barcode_scales = [10]

fixed_exp1 = dict(
    backbone_length     = BACKBONE_LEN,
    backbone_error_rate = ERROR_RATE,
)

rows_exp1 = []
for n in barcode_scales:
    depth  = max(1, TOTAL_READS // n)
    kwargs = {**fixed_exp1, "num_barcodes": n, "read_depth": depth}
    out_dir = f"{OUT_ROOT}/exp1/{n}"
    print(f"  [exp1] num_barcodes={n}, read_depth={depth} ...")

    fasta, csv, fastq, barcodes, gt = generate_benchmark_data(out_dir, **kwargs)
    naive_fasta                     = generate_naive_fasta(out_dir, fasta, csv)

    la_t,  la_m,  la_bam  = benchmark_library_aligner(fasta, csv, fastq, out_dir, PRESET, KMER, W)
    lac_t, lac_m, lac_bam = benchmark_library_aligner(fasta, csv, fastq, out_dir, PRESET, KMER, W, no_cache=True)
    nv_t,  nv_m,  nv_sam  = benchmark_naive(naive_fasta, fastq, out_dir, PRESET, KMER, W)

    la_acc  = measure_accuracy_la(la_bam,  gt)
    lac_acc = measure_accuracy_la(lac_bam, gt)
    nv_acc  = measure_accuracy_naive(nv_sam, gt)

    rows_exp1.append(dict(
        num_barcodes        = n,
        read_depth          = depth,
        la_time             = la_t,   la_mem             = la_m,   la_accuracy        = la_acc,
        la_nocache_time     = lac_t,  la_nocache_mem     = lac_m,  la_nocache_accuracy= lac_acc,
        naive_time          = nv_t,   naive_mem          = nv_m,   naive_accuracy     = nv_acc,
    ))
    print(f"    cached:   {la_t:.1f}s {la_m:.0f}MB acc={la_acc:.3f}")
    print(f"    no-cache: {lac_t:.1f}s {lac_m:.0f}MB acc={lac_acc:.3f}")
    print(f"    naive:    {nv_t:.1f}s {nv_m:.0f}MB acc={nv_acc:.3f}")

df_exp1 = pd.DataFrame(rows_exp1)
df_exp1.to_csv(f"{OUT_ROOT}/exp1/results.csv", index=False)
print(df_exp1.to_string(index=False))
```

------------------------------------------------------------------------

# Experiment 2 --- Varying Read / Template Length

Number of barcodes and total reads are fixed; backbone length is swept from short (50 bp) to long-read realistic (3000 bp).

```{python exp2}
LENGTH_SCALES = [50, 200, 400, 1000, 2000, 3000]
N_BARCODES    = 250
READ_DEPTH    = 20

rows_exp2 = []
for bl in LENGTH_SCALES:
    out_dir = f"{OUT_ROOT}/exp2/{bl}"
    print(f"  [exp2] backbone_length={bl} ...")

    fasta, csv, fastq, barcodes, gt = generate_benchmark_data(
        out_dir,
        num_barcodes        = N_BARCODES,
        read_depth          = READ_DEPTH,
        backbone_length     = bl,
        backbone_error_rate = ERROR_RATE,
    )
    naive_fasta = generate_naive_fasta(out_dir, fasta, csv)

    la_t,  la_m,  la_bam  = benchmark_library_aligner(fasta, csv, fastq, out_dir, PRESET, KMER, W)
    lac_t, lac_m, lac_bam = benchmark_library_aligner(fasta, csv, fastq, out_dir, PRESET, KMER, W, no_cache=True)
    nv_t,  nv_m,  nv_sam  = benchmark_naive(naive_fasta, fastq, out_dir, PRESET, KMER, W)

    la_acc  = measure_accuracy_la(la_bam,  gt)
    lac_acc = measure_accuracy_la(lac_bam, gt)
    nv_acc  = measure_accuracy_naive(nv_sam, gt)

    read_len = bl * 2 + len(F5) + len(F3) + 15  # approx full resolved length
    rows_exp2.append(dict(
        backbone_length     = bl,
        approx_read_len     = read_len,
        la_time             = la_t,   la_mem             = la_m,   la_accuracy        = la_acc,
        la_nocache_time     = lac_t,  la_nocache_mem     = lac_m,  la_nocache_accuracy= lac_acc,
        naive_time          = nv_t,   naive_mem          = nv_m,   naive_accuracy     = nv_acc,
    ))
    print(f"    cached:   {la_t:.1f}s {la_m:.0f}MB acc={la_acc:.3f}")
    print(f"    no-cache: {lac_t:.1f}s {lac_m:.0f}MB acc={lac_acc:.3f}")
    print(f"    naive:    {nv_t:.1f}s {nv_m:.0f}MB acc={nv_acc:.3f}")

df_exp2 = pd.DataFrame(rows_exp2)
df_exp2.to_csv(f"{OUT_ROOT}/exp2/results.csv", index=False)
print(df_exp2.to_string(index=False))
```

------------------------------------------------------------------------

# Experiment 3 --- Varying Template Divergence

Number of barcodes, read depth, and backbone length are fixed; the per-base substitution rate between templates is swept from 0 (identical backbones) to 1.0 (completely random backbones). This tests how sequence similarity between references affects alignment accuracy and speed.

```{python exp3}
DIVERGENCE_SCALES = [0.0, 0.01, 0.05, 0.1, 0.25, 0.5, 1.0]
N_BARCODES_EXP3   = 500
READ_DEPTH_EXP3   = 2

rows_exp3 = []
for er in DIVERGENCE_SCALES:
    out_dir = f"{OUT_ROOT}/exp3/{er}"
    print(f"  [exp3] backbone_error_rate={er} ...")

    fasta, csv, fastq, barcodes, gt = generate_benchmark_data(
        out_dir,
        num_barcodes        = N_BARCODES_EXP3,
        read_depth          = READ_DEPTH_EXP3,
        backbone_length     = BACKBONE_LEN,
        backbone_error_rate = er,
    )
    naive_fasta = generate_naive_fasta(out_dir, fasta, csv)

    la_t,  la_m,  la_bam  = benchmark_library_aligner(fasta, csv, fastq, out_dir, PRESET, KMER, W)
    lac_t, lac_m, lac_bam = benchmark_library_aligner(fasta, csv, fastq, out_dir, PRESET, KMER, W, no_cache=True)
    nv_t,  nv_m,  nv_sam  = benchmark_naive(naive_fasta, fastq, out_dir, PRESET, KMER, W)

    la_acc  = measure_accuracy_la(la_bam,  gt)
    lac_acc = measure_accuracy_la(lac_bam, gt)
    nv_acc  = measure_accuracy_naive(nv_sam, gt)

    rows_exp3.append(dict(
        backbone_error_rate  = er,
        la_time              = la_t,   la_mem              = la_m,   la_accuracy         = la_acc,
        la_nocache_time      = lac_t,  la_nocache_mem      = lac_m,  la_nocache_accuracy = lac_acc,
        naive_time           = nv_t,   naive_mem           = nv_m,   naive_accuracy      = nv_acc,
    ))
    print(f"    cached:   {la_t:.1f}s {la_m:.0f}MB acc={la_acc:.3f}")
    print(f"    no-cache: {lac_t:.1f}s {lac_m:.0f}MB acc={lac_acc:.3f}")
    print(f"    naive:    {nv_t:.1f}s {nv_m:.0f}MB acc={nv_acc:.3f}")

df_exp3 = pd.DataFrame(rows_exp3)
df_exp3.to_csv(f"{OUT_ROOT}/exp3/results.csv", index=False)
print(df_exp3.to_string(index=False))
```

------------------------------------------------------------------------

# Plots

```{r load-results, message=FALSE, warning=FALSE}
library(tidyverse)
library(patchwork)

theme_set(
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold"),
    legend.position = "bottom"
  )
)

COLS <- c(
  "library-aligner (cached)"   = "#2166ac",
  "library-aligner (no-cache)" = "#74add1",
  "naive minimap2"             = "#d73027"
)

pivot_results <- function(df, x_var) {
  df |>
    select(all_of(x_var),
           starts_with("la_"), starts_with("naive_")) |>
    pivot_longer(
      -all_of(x_var),
      names_to  = c("method", ".value"),
      names_pattern = "^(la_nocache|la|naive)_(.*)"
    ) |>
    mutate(method = recode(method,
      la        = "library-aligner (cached)",
      la_nocache= "library-aligner (no-cache)",
      naive     = "naive minimap2"
    ))
}

exp1 <- read_csv("/tmp/la_bench/exp1/results.csv", show_col_types = FALSE)
exp2 <- read_csv("/tmp/la_bench/exp2/results.csv", show_col_types = FALSE)
exp3 <- read_csv("/tmp/la_bench/exp3/results.csv", show_col_types = FALSE)
```

## Experiment 1 --- Number of Barcodes

```{r plot-exp1, fig.width=12, fig.height=10}
d1 <- pivot_results(exp1, "num_barcodes")

COLS <- c(
  "library-aligner (cached)"   = "#2166ac",
  "library-aligner (no-cache)" = "#f46d43",
  "naive minimap2"             = "#1a9641"
)

p1a <- ggplot(d1, aes(num_barcodes, time, colour = method)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_colour_manual(values = COLS) +
  scale_x_log10() +
  labs(title = "Run time vs. library size",
       x = "Number of barcodes (log scale)", y = "Wall time (s)",
       colour = NULL) +
  ylim(0, NA)

p1b <- ggplot(d1, aes(num_barcodes, mem, colour = method)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_colour_manual(values = COLS) +
  scale_x_log10() +
  labs(title = "Peak memory vs. library size",
       x = "Number of barcodes (log scale)", y = "Peak RSS (MB)",
       colour = NULL) +
  ylim(0, NA)

p1c <- ggplot(d1, aes(num_barcodes, accuracy, colour = method)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_colour_manual(values = COLS) +
  scale_x_log10() +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(title = "Mapping accuracy vs. library size",
       x = "Number of barcodes (log scale)", y = "Accuracy",
       colour = NULL) +
  ylim(0, NA)

library(cowplot)



legend_plot <- ggplot(d1, aes(num_barcodes, time, colour = method)) +
  geom_line() +
  scale_colour_manual(values = COLS, name = "Method") +
  theme(legend.position = "right")

legend <- cowplot::get_plot_component(legend_plot, "guide-box-right")

plots <- plot_grid(
  p1a + theme(legend.position = "none"),
  p1b + theme(legend.position = "none"),
  p1c + theme(legend.position = "none"),
  ncol = 1
)
plot_grid(plots, legend, ncol = 2, rel_widths = c(1, 0.3))


```

## Experiment 2 --- Read / Template Length

```{r plot-exp2, fig.width=12, fig.height=10}
d2 <- pivot_results(exp2, "backbone_length")

p2a <- ggplot(d2, aes(backbone_length, time, colour = method)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_colour_manual(values = COLS) +
  labs(title = "Run time vs. template length",
       x = "Backbone length (bp)", y = "Wall time (s)", colour = NULL) +
  ylim(0, NA)

p2b <- ggplot(d2, aes(backbone_length, mem, colour = method)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_colour_manual(values = COLS) +
  labs(title = "Peak memory vs. template length",
       x = "Backbone length (bp)", y = "Peak RSS (MB)", colour = NULL) +
  ylim(0, NA)

p2c <- ggplot(d2, aes(backbone_length, accuracy, colour = method)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_colour_manual(values = COLS) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(title = "Mapping accuracy vs. template length",
       x = "Backbone length (bp)", y = "Accuracy", colour = NULL) +
  ylim(0, NA)

legend_plot2 <- ggplot(d2, aes(backbone_length, time, colour = method)) +
  geom_line() +
  scale_colour_manual(values = COLS, name = "Method") +
  theme(legend.position = "right")
legend2 <- cowplot::get_plot_component(legend_plot2, "guide-box-right")
plots2 <- plot_grid(
  p2a + theme(legend.position = "none"),
  p2b + theme(legend.position = "none"),
  p2c + theme(legend.position = "none"),
  ncol = 1
)
plot_grid(plots2, legend2, ncol = 2, rel_widths = c(1, 0.3))
```

## Experiment 3 --- Template Divergence

```{r plot-exp3, fig.width=12, fig.height=10}
d3 <- pivot_results(exp3, "backbone_error_rate")

p3a <- ggplot(d3, aes(backbone_error_rate, time, colour = method)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_colour_manual(values = COLS) +
  scale_x_continuous(labels = scales::percent) +
  labs(title = "Run time vs. template divergence",
       x = "Per-base substitution rate", y = "Wall time (s)", colour = NULL) +
  ylim(0, NA)

p3b <- ggplot(d3, aes(backbone_error_rate, mem, colour = method)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_colour_manual(values = COLS) +
  scale_x_continuous(labels = scales::percent) +
  labs(title = "Peak memory vs. template divergence",
       x = "Per-base substitution rate", y = "Peak RSS (MB)", colour = NULL) +
  ylim(0, NA)

p3c <- ggplot(d3, aes(backbone_error_rate, accuracy, colour = method)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_colour_manual(values = COLS) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(title = "Mapping accuracy vs. template divergence",
       x = "Per-base substitution rate", y = "Accuracy", colour = NULL) +
  ylim(0, NA)

legend_plot3 <- ggplot(d3, aes(backbone_error_rate, time, colour = method)) +
  geom_line() +
  scale_colour_manual(values = COLS, name = "Method") +
  theme(legend.position = "right")
legend3 <- cowplot::get_plot_component(legend_plot3, "guide-box-right")
plots3 <- plot_grid(
  p3a + theme(legend.position = "none"),
  p3b + theme(legend.position = "none"),
  p3c + theme(legend.position = "none"),
  ncol = 1
)
plot_grid(plots3, legend3, ncol = 2, rel_widths = c(1, 0.3))
```

------------------------------------------------------------------------

# Session Info

```{r session}
sessionInfo()
```

```{python versions}
import mappy, pysam, edlib
print(f"mappy  {mappy.__version__}")
print(f"pysam  {pysam.__version__}")
```
